<!-- Change the example data to use HousePrices data set-->

# Simple linear regression

Regression analysis is a statistical procedure that allows us to **model relationships between variables** so that the causal relationship between variables is defined. It can be considered as the main statistical technique used in economics, i.e. econometrics.  There is a large **variety of regression models**, depending on estimation method, model specification and assumed distributions. This section introduces the most basic of these, the simple linear regression model, i.e. ***ordinary least squares (OLS)*** with one independent variable.

## Ordinary least squares

See @navarro_learning_2018 section 12.3 for an introduction.

### Model specification

Regression models represent how one or more more predictor variables have an effect on a response variable. The most simple regression model has **only one predictor**, which is why it is called "simple linear regression" (SLS). An SLS model is mathematically expressed as follows:

$$\hat{y} = \beta_0 + \beta_1 x + \varepsilon,$$

where

- $\hat{y}$ is dependent or explained variable, **response** or regressand, 
- $\beta_0$ is the intercept or constant, 
- $\beta_1$ is a coefficient of $x$, 
- $x$ is independent or explanatory variable, **predictor** or regressor, and
- $\varepsilon$ is the model error or residual.

The values of $x$ and $y$ come from our data, whereas the $\beta$ coefficients need to be calculated from this data. The $\varepsilon$ is estimated using both, the $\beta$ coefficients and data.

### Calculation of an SLS model

The underlying idea behind OLS regression is the **minimization of squared residuals**, thus the name "least squares". We simply find the coefficients $\beta_0$ and $\beta_1$ that give us the smallest possible sum of differences between observed and predicted values of the response variable $y$. 

Model parameters are calculated directly (unlike maximum likelihood estimation that is based on iterations). To estimate the model $\hat{y} = \beta_0 + \beta_1 x + \varepsilon$ for variables $x$ and $y$ we estimate the parameters $\hat{\beta_1}$ and $\hat{\beta_0}$ as follows:

1. $\beta_1 = \frac{\sum{x_{i} y_{i}} - \frac{1}{n} \sum{x_{i}}\sum{y_{i}}}{\sum{x_{i}^{2}} - \frac{1}{n} (\sum{x_{i}})^{2}} = \frac{Cov[x,y]} {Var[x]},$

2. $\beta_0 = \overline{y} - \beta_1 \overline{x}.$

For a simple model $y = \beta_1 x + \varepsilon$ we can simply use matrix algebra on values of $x$ and $y$ to find $\hat{\beta_1}$:

$$\hat{\beta_1} = (X^{T} X)^{-1} X^{T} Y,$$

where $X$ is the column matrix of predictor and $Y$ the column matrix of response.

See @navarro_learning_2018 section 12.4 for illustrations of the idea.

In Jamovi: `Regression > Linear regression     `.  
In R: `lm(y ~ x, data)` or `summary(lm(y ~ x, data))`

## Elements of a regression model

### Intercept

In the equation this is the $\beta_0$ and often also referred to as the *constant*. Intercept is the value of $y$ where regression line crosses the Y-axis, so **intercept is the value of $y$ when $x$ is zero** ( $y|x=0$ ). Intercept does not need to be theoretically valid but it sometimes is. In particular, the intercept is not relevant when the range of our observed values of $x$ does not cover 0.

For example, the model below has an intercept.

``` {r, echo = F}
calves <- agridat::urquhart.feedlot
names(calves)[names(calves) %in% c('weight1', 'weight2')] <- c('initialwt', 'finalwt')
wtMod <- lm(finalwt ~ initialwt, calves)
```

$$finalwt_{i} = `r round(wtMod$coef[[1]], 3)` + `r round(wtMod$coef[[2]], 3)` * initialwt_{i} + \varepsilon_{i}$$

> When `initialwt` is 0 units, then the value of `finalwt` is `r round(wtMod$coef[[1]], 3)` units.

### Coefficient(s)

In the equation the $\beta_1$ represents coefficient of $x$. It indicates **by how many units $y$ increases when $x$ increases by one unit**. 

We usually estimate models on a sample and wish to make generalizations or inferences about population. It is thus necessary to test whether or not coefficients are different from 0. This is done by calculating t-statistic from coefficient and standard error. **Coefficients are only relevant if their difference from 0 is statistically significant**. If a coefficient is not statistically significant, then we would conclude that the predictor has no effect on the response.

The interpretation of $\beta$ coefficients is different for interval or ratio (numeric) predictors and ordinal or nominal (categorical) predictors.

``` {r, echo = F}
calves <- agridat::urquhart.feedlot
names(calves)[names(calves) %in% c('weight1', 'weight2')] <- c('initialwt', 'finalwt')
wtMod <- lm(finalwt ~ initialwt, calves)
calves$diet <- factor(calves$diet, levels = c('Low', 'Medium', 'High'))
wtModCat <- lm(finalwt ~ diet, calves)
```

In the following example we estimate the effect of a ratio (numeric) predictor  `initialwt` on response `finalwt`.

$$finalwt_{i} = `r round(wtMod$coef[[1]], 3)` + `r round(wtMod$coef[[2]], 3)` * initialwt_{i} + \varepsilon_{i}$$

> When `initialwt` increases by 1 unit, then `finalwt` increases by **`r round(wtMod$coef[[2]], 3)`** units.

Suppose that a categorical variable called `diet` has the following levels: *`r levels(calves$diet)`*. Then in regression analysis the first level (*`r levels(calves$diet)[1]`*) is considered as the base level. All other factor values are compared against this base value.

$$finalwt_{i} = `r round(wtModCat$coef[[1]], 3)` + `r round(wtModCat$coef[[2]], 3)`dietMedium_{i} + `r round(wtModCat$coef[[3]], 3)`dietHigh_{i} + \varepsilon_{i}$$

> When `diet` is *`r levels(calves$diet)[1]`*, then `finalwt` is **`r round(wtModCat$coef[[1]], 3)`**.

> When `diet` is *Medium*, then `finalwt` is higher by **`r abs(round(wtModCat$coef[[2]], 3))`** units compared to when `diet` is *`r levels(calves$diet)[1]`*, i.e. `r round(wtModCat$coef[[1]] + wtModCat$coef[[2]], 3)`.

Suppose now that the coefficient for `dietHigh` is not statistically significant. When `diet` is *High*, `finalwt` is no different compared to when `diet` is *`r levels(calves$diet)[1]`*. If it was statistically significant, we could say that when `diet` is *High*, then `finalwt` is higher by **`r abs(round(wtModCat$coef[[3]], 3))`** units compared to when `diet` is *`r levels(calves$diet)[1]`*.

### Fitted values

These are the estimated values of response ( $\hat{y}$ ) that are calculated by applying the model for $x$ of every observation in the data. In other words, **fitted values are predictions of $y$ based on a model**. After we have estimated the $\beta_0$ and $\beta_1$ we can calculate fitted values for each value of $x$ as follows: $\hat{y} = \beta_0 + \beta_1 x$.

### Residuals

Residuals are model errors, represented by the $\varepsilon$ in the equation $\hat{y} = \beta_0 + \beta_1 x + \varepsilon$. Residuals are the **difference between observed and fitted (predicted) values of response variable**. So we can find residuals for each observation after we have estimated $\beta_0$ and $\beta_1$ using values of $x$ and $y$ as follows: $\varepsilon = y - \beta_0 -  \beta_1 x$.

We use residuals to evaluate how well model fits data. Large residuals would mean that fitted values are different from observed values, which indicates that model does not fit data well.

### The $R^2$ 

This is a way to measure ***goodness of fit*** of a model, i.e. how well model fits data. It is sometimes called *coefficient of determination*. $R^2$ expresses the **part of variation in response variable that is explained by model**:

$$R^{2} = \frac{ESS}{TSS} = 1 - \frac{RSS}{TSS},$$

where the elements are defined a follows:

- **e**xplained sum of squares, $ESS$; 
  $\sum_{i = 1}^{n} (\hat{y}_{i} - \bar{y})^2$
- **r**esidual sum of squares, $RSS$; 
  $\sum_{i = 1}^{n} (y_{i} - \hat{y}_{i})^2$
- **t**otal sum of squares, $TSS$; 
  $\sum_{i = 1}^{n} (y_{i} - \bar{y})^2.$

In above equations the following notation is used: 

- $\hat{y}_{i}$ are fitted values, 
- $\overline{y}$ is the mean value of $y$, and
- $y_{i}$ are the observed values of $y$.

Simply put, the $R^2$ measures how much better is model at explaining the variance of $y$ compared to how well just the mean of $y$ explains variance of $y$.

### The adjusted- $R^2$ 

Each additional predictor likely improves the model fit. So the more predictors we add, the more the model explains variance of $y$. So $R^2$ can be inflated just by adding more predictors to the model even if the predictors are not very meaningful. To **penalize a model for the number of predictors ( $k$ )** while considering the number of observations ( $n$ ), the adjusted $R^{2}$ can also be used, particularly for model comparison:

$$\overline{R^{2}} = 1 - \frac{RSS/(n-k)}{TSS/(n-1)}$$

## Assumptions and diagnostics

See @navarro_learning_2018 section 12.9 for an explanation of assumptions and section 12.10 on model checking.

After the estimation of a regression model it should be diagnosed to make sure that it meets at least the following assumptions:

- residuals are **normally distributed**, and
- residuals have **equal variance**, i.e. variance of residuals does not depend on the values of $x$

In Jamovi: `Regression > Linear regression > Assumption Checks    `.  
In R: `plot(lm(formula, data))`

### Gauss-Markov assumptions

In addition to the practical considerations outlined above, a  theoretical way of expressing the assumptions of OLS is via the Gauss–Markov theorem. It posits the following assumptions:

- linear in parameters  
  $Y = \beta_0 + \beta_1 x + \varepsilon$
- expected error is zero  
  $E(\varepsilon) = 0$
- homoscedasticity  
  $var(\varepsilon) = E(\varepsilon^{2})$
- no autocorrelation  
  $cov(\varepsilon_{i}, \varepsilon_{j}) = 0, i \neq j$
- independence of predictor(s) and residuals  
  $cov(x,\varepsilon) = 0$

If these are true, then the model is the *best linear unbiased estimator* (BLUE).