---
title: Cluster analysis
subtitle: Research methods
author: Jüri Lillemets
date: "`r Sys.Date()`"
output_yaml: _output.yml
editor_options: 
  chunk_output_type: console
---

``` {r setup, include = F}
# Settings
knitr::opts_chunk$set(echo = F, warning = F, message = F, dpi = 200, fig.height = 4)
# Load packages
library('magrittr');library('knitr')
library('cluster')
# Options
options(scipen = F, digits = 3)
# Set colors
Col <- c(red = '#e6457a', green = '#b0e645', blue = '#45cbe6')
```

class: center middle clean

# How to group objects?

---

class: center middle inverse

# What is clustering?

---

## Clustering and classification

In classification we already know the classes and use that information to determine classification rules.

In clustering we do not have any information on possible existing classes and we can not compare clusters to classes.

---

## Why objects?

We can cluster both observations and variables. 

That's why we refer to **objects** as the phenomena to be assigned to clusters.

---

## Standardization

Variables with higher variances have a higher influence on how the objects are clustered. 

If this is not desired, variables should be standardized.

---

Why should we standardize?

``` {r}
data(cloth, package = 'boot')
par(bty = 'n', family = 'RobotoCondensed', mfrow = 1:2)
plot(cloth, main = "Raw data", 
     xlab = "The length of the roll of cloth", 
     ylab = "The number of flaws found in the roll", 
     pch = 20, asp = 1)
plot(scale(cloth), main = "Scaled data", 
     xlab = "The length of the roll of cloth", 
     ylab = "The number of flaws found in the roll", 
     pch = 20, asp = 1)
```

---

Clustering can be applied in practice for various purposes.

Marketing and sales - find homogeneous groups of customers so that promotional campaigns could be addresses more accurately and thus more efficiently.

Medicine - cluster patients with similar symptoms or predispositions for treatment or discovery of risk.

Finance - categorize enterprises into different types based on some financial or other characteristics.

Biology - assign plants to species depending on characteristics they share.

--

We can cluster anything we want.

---

## Methods

**Model-based** methods where we assume that objects follow mixture of distributions

**Distance-based** methods where we use distances between objects  
  - **hierarchical** clustering methods
  - **partitioning** methods
  
---

## Distance measures

For distance-based measures we use distances between objects to determine how to assign them into clusters. We look at the most common measures for continuous variables, the Euclidean and Manhattan distances: 

$$d_{Euclidean}(x_i,x_j) = [\sum^p_{k=1}(x_{ik} - x_{jk})^2]^{1/2},$$

$$d_{Manhattan}(x_i,x_j) = \sum^p_{k=1} |x_{ik} - x_{jk}|.$$

--

Distance measures for ordinal and nominal variables also exist but are not explained here.

---

Example data on the number of flaws in cloth for 32 pieces of cloth.

``` {r}
# Visualizatin of Euclidean and Manhattan distances
x <- cloth[c(6,27), 1]; y <- cloth[c(6,27), 2]
par(bty = 'n', family = 'RobotoCondensed', mfrow = 1:2)
# Euclidean
plot(cloth, main = "Euclidean distance", 
     xlab = "The length of the roll of cloth", 
     ylab = "The number of flaws found in the roll", pch = 20)
lines(x[1:2], y[1:2], col = Col['red'], lwd = 2)
# Manhattan
plot(cloth, main = "Manhattan distance", 
     xlab = "The length of the roll of cloth", 
     ylab = "The number of flaws found in the roll", pch = 20)
lines(c(x[1], x[2]), c(y[1], y[1]), col = Col['red'], lwd = 2)
lines(c(x[2], x[2]), c(y[1], y[2]), col = Col['red'], lwd = 2)
```

---

class: center middle inverse

# Hierarchical clustering

---

Clusters are constructed **incrementally**. This can be done

- **divisive**ly ("top-down") where we start from most different objects or
- **agglomerative**ly ("bottom-up") where we start from most similar objects.

We will explore the agglomerative hierarchical clustering (Agnes - agglomerative nesting).

---

The example data we use contains on 908 measurements on three hawk species.

.compact[
``` {r}
data('Hawks', package = 'Stat2Data')
tail(Hawks[10:19], 10) %>% kable
```
]

---

We will use the following variables to cluster the birds.

- `Wing`	Length (in mm) of primary wing feather from tip to wrist it attaches to
- `Weight`	Body weight (in gm)
- `Culmen`	Length (in mm) of the upper bill from the tip to where it bumps into the fleshy part of the bird
- `Hallux`	Length (in mm) of the killing talon
- `Tail`	Measurement (in mm) related to the length of the tail (invented at the MacBride Raptor Center)
- `StandardTail`	Standard measurement of tail length (in mm)
- `Tarsus`	Length of the basic foot bone (in mm)
- `WingPitFat`	Amount of fat in the wing pit
- `KeelFat`	Amount of fat on the breastbone (measured by feel
- `Crop`	Amount of material in the crop, coded from 1=full to 0=empty

---


## Process

Agglomerative hierarchical clustering has the following steps.

1. Initial number of clusters is $n$, so each cluster contains one object. 
2. Calculate distance matrix $D$ that expresses pairwise distances between clusters (objects).
3. Find the smallest distance and merge the objects with smallest distance into a single cluster.
4. Calculate a new distance matrix $D$ that now includes distance between the new cluster and all other clusters using a linkage method (see below).
5. Repeat the previous two steps until all objects are in a single cluster.

---

Let's illustrate the process with two variables and 5 hawks.

``` {r}
Sample <- Hawks[6:11, c(10,14)]
par(bty = 'n', family = 'RobotoCondensed', mar = c(4,4,0,0))
plot(Sample, 
     xlab = "Wing length, mm", ylab = "Tail length, mm", 
     pch = 20, cex = 2,  asp = 1)
text(Sample, rownames(Sample), 0, 2)
```

---

Here's the respective initial Euclidean distance matrix $D$.

``` {r}
dist(Sample) %>% as.matrix %>% kable
```

---

exclude: true

What if we have more than two variables? For example three?

``` {r, eval = F}
par(bty = 'n', family = 'RobotoCondensed', mfrow = 1:2)
rgl::plot3d(Hawks[6:11, c(10,11,14)], 
            col = Col['red'], size = 2, type = 's', pch = 20, asp = 1)
rgl::rglwidget(width = 400)
```

---

## Linkage methods

How to calculate distance between cluster and a point?

- single linkage: $d_{IJ,k} = min(d_{i,K}, d_{j,K})$;
- complete linkage: $d_{IJ,k} = max(d_{I,K}, d_{J,K})$; 
- average linkage: $d_{IJ,k} = \sum_{i \in IJ} \sum_{k \in K} d_{ik} / (n_{ij}n_k)$; 
- Ward's method: compares the within-cluster and between-cluster squared distances.

---

How to think about single, complete and average linkage?

``` {r}
par(bty = 'n', family = 'RobotoCondensed')
plot(Sample, 
     xlab = "Wing length, mm", ylab = "Tail length, mm", 
     pch = 20, cex = 2,  asp = 1)
text(Sample, rownames(Sample), 0, 2)
```

???
Draw linkage results.

---

## Dendrogram

The clusters in case of hierarchical clustering are estimated incrementally, resulting in a nested structure. This tree-shaped structure can be visualized by a **dendrogram**.

---

``` {r} 
par(bty = 'n', family = 'RobotoCondensed', mfrow = 1:2)
i <- 'ward.D2'
dist(Sample) %>% hclust(method = i) %>% plot(main = paste("Linkage:", toupper(i)))
par(bty = 'n', family = 'RobotoCondensed')
plot(Sample, 
     xlab = "Wing length, mm", ylab = "Tail length, mm", 
     pch = 20, cex = 2,  asp = 1)
text(Sample, rownames(Sample), 0, 2)
```

---

Dendrogram allows us to also illustrate differences between linkage methods.

``` {r} 
par(bty = 'n', family = 'RobotoCondensed', mfrow = c(1,4))
for (i in c('single', 'complete', 'average', 'ward.D2')) {
  dist(Sample) %>% hclust(method = i) %>% plot(main = paste("Linkage:", toupper(i)))
}
```

---

To determine the clusters we can choose either the height of cut or number of clusters. 

``` {r} 
par(bty = 'n', family = 'RobotoCondensed', mar = c(0,4,4,0))
Clust <- dist(Hawks[1:100, 10:19]) %>% hclust(method = 'ward.D2')
plot(Clust, cex = .5)
```

---

If we would cut at 1000 then we would obtain `r max(cutree(Clust, h = 1000))` clusters.

``` {r} 
par(bty = 'n', family = 'RobotoCondensed', mar = c(0,4,4,0))
Clust <- dist(Hawks[1:100, 10:19]) %>% hclust(method = 'ward.D2')
plot(Clust, cex = .5)
abline(h = 1000, col = Col['red'])
```

--

> At what height would we have to cut if we wished to obtain 3 clusters?

---

Actually, the species for each Hawk is already determined. How does it coincide with our clusters?

``` {r}
table(dist(Hawks[10:19]) %>% hclust(method = 'ward.D2') %>% cutree(3), 
      Hawks$Species) %>% 
  kable
```

---

class: center middle inverse

# K-means clustering

---

Clusters are constructed by **partitioning** objects **iteratively**.

The goal is to partition objects $x$ into $K$ clusters so that distances between objects within cluster are small compared to distances to points outside the cluster. 

We can achieve this by assigning each object to the closest **centroid**, i.e. cluster mean.

---

We thus need optimal cluster means. The optimal mean vector $\bar x_1,  ... \bar x_K$ can be found by minimizing the following function:

$$ESS = \sum^K_{k = 1} \sum_{c(i)=k} (x_i - \bar  x_k)^T(x_i - \bar x_k),$$

where $c(i)$ is the cluster containing $x_i$. 

--

An alternative is **K-medoids clustering** in which case the centroids are not some mean values but represented by actual objects.

???
Number of clusters has to be defined before clustering.
We attempt to minimize the sum of distances within all clusters

---

Let's attempt to cluster 392 vehicles.

``` {r}
data('Auto', package = 'ISLR')
head(Auto[, 1:7], 4) %>% kable
```

--

We should scale the variables.

``` {r}
Oie <- scale(Auto[, 1:7])
head(Oie, 4) %>% kable
```

---

We have 7 variables, so 7 dimensions. Let's use PCs to represent the data.

``` {r, fig.height = 4.5}
par(family = 'RobotoCondensed')
Auto[, 1:7] %>% pairs(pch = 20, cex = .5)
```

---

What do we mean by "partitioning"?

``` {r}
PCA <- prcomp(Oie)
PCs <- PCA$x[, 1:2]
par(bty = 'n', family = 'RobotoCondensed', pty = 's', mar = c(4,4,0,0))
plot(PCs, pch = 20)
```

---

## Gap statistic

The Gap statistic is a technique used to determine the optimal number of clusters.

The measure compares the sum of average distances within cluster to the same sum obtained from uniformly distributed data. 

The optimal number of clusters is at the value of the Gap statistic $k$ where $k + se(k)$ is higher or equal to the estimate $k$ of next number of clusters.

---

``` {r}
kmFun <- function(x, k) kmeans(x, k, iter.max = 50, nstart = 10)
kmG <- clusGap(Oie, kmFun, K.max = 8, B = 10)
par(bty = 'n', family = 'RobotoCondensed')
plot(kmG)
```

> How many clusters should we estimate?

---

## Process

K-means clustering involves the following steps:

1. We start with a distance matrix $D$ based on 
   - random assignment of objects to $K$ clusters with cluster means, or
   - some (random) cluster means.
2. Calculate squared Euclidean distance between each object and each cluster mean. Reassign each item to its nearest cluster mean, resulting in decreased $ESS$.
3. Update cluster means.
4. Repeat the previous two steps until objects can not be reassigned, so each object is closest to its own cluster mean.

---

Can we see any clusters if we summarize the data into two principal components?

``` {r}
par(bty = 'n', family = 'RobotoCondensed', pty = 's', mar = c(4,4,0,0))
plot(PCs, pch = 20)
```

> How many clusters would you distinguish?

---

exclude: true

``` {r, eval = F}
par(bty = 'n', family = 'RobotoCondensed')
rgl::plot3d(prcomp(Oie)$x[, 1:3], 
            col = Col['red'], size = 1, type = 's', pch = 20, asp = 1)
```

---

Let's esimate 4 clusters as suggested by the Gap statistic. Depending on the centers we choose, we obtain different clusters.

``` {r}
par(bty = 'n', family = 'RobotoCondensed', pty = 's', mfrow = c(2,3), 
    mar = c(4,4,0,0))
replicate(6, {
  Kmeans <- kmeans(PCs, 4, iter.max = 1, nstart = 1)
  plot(PCs, pch = 20, col = rainbow(4)[Kmeans$cluster])
  Kmeans$centers %>% points(col = 'black', pch = 4, lwd = 2)
  }) %>% invisible # To hide printing from replicate
```

???
Cluster number is arbitrary.

---

## Cluster plot

When we apply K-means clustering we do not obtain a nested structure and therefore can not express the clustering as a dendrogram. 

We can create a plot that shows the locations of objects and cluster centroids on a two-dimensional plot.

The dimensions can be found via PCA or multidimensional scaling.

---

Here is the final estimate if we run 10 iterations and 10 different starting values.

``` {r}
par(bty = 'n', family = 'RobotoCondensed', pty = 's', mar = c(4,4,0,0))
Kmeans <- kmeans(PCs, 4, iter.max = 10, nstart = 10)
plot(PCs, pch = 20, col = rainbow(4)[Kmeans$cluster])
Kmeans$centers %>% points(col = 'black', pch = 4, lwd = 2)
```

---

class: center middle inverse

# How to use clusters?

---

Save clusters as a new variable.

.compact[
``` {r}
cbind(Auto, Cluster = Kmeans$cluster) %>% head %>% kable
```
]

---

You can describe clusters using descriptive statistics, e.g. mean values as below

.compact[
``` {r}
aggregate(Auto[, 1:7], by = list(Cluster = Kmeans$cluster), FUN = mean) %>% kable
```
]

---

Or create some visualizations.

``` {r, fig.height = 4.5}
par(bty = 'n', family = 'RobotoCondensed', mfrow = c(2,3), mar = c(4,4,2,2))
for (i in c(1,3:7)) {
  boxplot(Auto[, i] ~ Kmeans$cluster, xlab = "Cluster", ylab = names(Auto)[i], 
          col = rainbow(4)[unique(Kmeans$cluster)])
}
```

---

class: inverse

???
Explain starting values and iterations.
Better color scale
